A Data-Driven Framework for Neural Field Modelling

### 2010-11-18 ltx/first_order_framework/LaTeX_Manuscript.tex
final commit before moving to different neuroimage cls

\subsection{Integro-Difference Equation Neural Field Model}
The combination of modelling techniques in this paper leads to a large amount of notation, so a reference of the symbols used is provided in Table~\ref{tab:Notation}. The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at time $t$ and position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations is described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right) \, dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}, 
\end{equation}
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations at positions $\mathbf{r}$ and $\mathbf{r}'$ are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\, d\mathbf{r}'}, 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' or ``wizard hat'' function, which describes local excitation and surround inhibition. An example of the Mexican hat connectivity kernel used in this paper is shown in Figure~\ref{fig:Figure1}. In an attempt model longer range connectivity, To demonstrate flexibility in the estimation algorithm and to model larger cortical regions, a third component of the kernel is introduced describing weak longer-range excitation. The exact shape of this kernel is assumed to vary across subjects, and hence needs to be inferred from subject-specific data.

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
To arrive at the final form of the model, we shall express the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation}
where $\textrm{D}=d/dt + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretise time using a first-order Euler method (see the supplementary (\dean{Appendix}) material for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean spatial Gaussian process with covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}. The disturbance is added to account for model uncertainty and unmodelled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

@@

When considering the measurement of a point source in a conductive homogeneous field measured by a point sensor, the observation kernel will theoretically follow a Laplacian like shape, with an exponential decay in the amplitude of a measurement with an increasing distance between the source and sensor~\cite{Jackson1999}. However, when considering a source or sensor with a spatial extent, the peak of the observation kernel becomes smoothed and is more similar to a Gaussian shape~\cite{Jackson1999}. An experiment supporting this was performed and the results are shown in \dean{Appendix?}. Therefore, in this study the observation kernel, $m(\mathbf{r}-\mathbf{r}')$, that governs the sensor pick-up geometry and is defined by 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)},
\end{equation}
where $\sigma_m$ sets the kernel width.

>>>>

\subsection{Integro-Difference Equation Neural Field Model}
The combination of modelling techniques in this paper leads to a large amount of notation, so a reference of the symbols used is provided in Table~\ref{tab:Notation}. The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at time $t$ and position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations is described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right) \, dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}, 
\end{equation}
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations at positions $\mathbf{r}$ and $\mathbf{r}'$ are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\, d\mathbf{r}'}, 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' or ``wizard hat'' function, which describes local excitation and surround inhibition. An example of the Mexican hat connectivity kernel used in this paper is shown in Figure~\ref{fig:Figure1}. To demonstrate flexibility in the estimation algorithm and to model larger cortical regions, a third component of the kernel is introduced describing weak longer-range excitation. The exact shape of this kernel is assumed to vary across subjects, and hence needs to be inferred from subject-specific data.

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
Now we shall express the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation}
where $\textrm{D}=d/dt + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretise time using a first-order Euler method (see the supplementary (\dean{Appendix}) material for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean spatial Gaussian process with covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}. The disturbance is added to account for model uncertainty and unmodelled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

@@

When considering the measurement of a point source in a conductive homogeneous field measured by a point sensor, the observation kernel will theoretically follow a Laplacian like shape, with an exponential decay in the amplitude of a measurement with an increasing distance between the source and sensor~\cite{Jackson1999}. However, when considering a source or sensor with a spatial extent, the peak of the observation kernel becomes smoothed and is more similar to a Gaussian shape~\cite{Jackson1999}. An experiment supporting this was performed and the results are shown in \dean{Appendix?}. Therefore, in this study the observation kernel, $m(\mathbf{r}-\mathbf{r}')$, that governs the sensor pick-up geometry is defined by 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)},
\end{equation}
where $\sigma_m$ sets the kernel width.
